Follow-up Samples

### 2017-01-18 submission/manuscript_outline_20161024.tex
streamline -discussion

Interestingly, when only previously associated cancer bacteria were
investigated only 2 out of the 4 had a significant decrease in relative
abundance between initial and follow up for the carcinoma group and 0
out of the 4 were significant for the adenoma group. This data suggests
that the changes that may be important in the bacterial microbiome after
surgery are not necessarily any one specific bacterium, a clear
depression of CRC related bacteria, nor any addition of new or depressed
bacterium but rather that a shift of the bacterial community as a whole
occurs with the existing members of this microbiome.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial and follow up samples in order to confirm that the community
was what was responding or changing due to surgery. We found that the
OTUs that made up the most important variables to the model
overwhelmingly belonged to comensal bacteria. With only the lesion model
having a single OTU from previously associated cancer bacteria
(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
and FIT both models (lesion and initial and follow up) had a
significantly decreased positive probability of either lesion or being
an initial sample on follow up {[}Figure 4 \& S4{]}. Confirming the
importance of the changes of comensal bacteria to these classifications
was that a total of 14 OTUs were common to both models and the vast
majority belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. One potential explanation is that even
with surgery and a shift of the bacterial community these specific
bacteria still persisted within the colon. An alternative reason would
be that even if they were reduced in relative abundance they were still
present. Another potential explanation is that they were not present in
enough individuals to be able to classify those with and without disease
with a high degree of accuracy. It is possible that our Random Forest
models were able to get the same information from measures such as FIT
or other OTUs. Finally, it is also possible that all three of these
potential explanations could have played a role. Regardless, our
observations would suggest that an individual's resident bacteria have a
large role to play in disease initiation and could change in a way that
allows predictive models to lower the positive probability of a lesion
after surgery {[}Figure 4{]}. It should be stated that our study does
not argue against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that the models do not utilize these
specific bacteria for classification purposes (lesion or initial and
follow up). In fact, it is possible that these CRC associated bacteria
are important in the transition from adenoma to carcinoma and would be
one explanation as to why in our data we not only see high initial
relative abundances, in certain individuals, but also large decreases in
relative abundance in those with carcinoma but not in those with adenoma
after surgery {[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model to have kept a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the fecal DNA test
{[}20{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of overfitting or having a model
that may not be immediately extrapolateable to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur.

Another interesting outcome was that Within figure 3 the initial and
follow up model showed better test AUC results then the training set
AUC. This may have occured because the training AUC that was determined
from 20 repeated 10 fold cross validation removed samples at random and
did not take into account that they were matched samples. Another
potential reason is that the model itself may be overfit since the total
number of samples was not that large. However, the lesion model did not
suffer from these discrepencies and similar conclusions can be drawn
solely from this model. Regardless, further independent studies will
need to be carried out to verify our findings since not only are we
dealing with feces, which could be very different than the communities
present on the actual tissue, but also are dealing with correlations
that may not be representative of the true pathogensis of disease.

>>>>

Interestingly, when only previously associated cancer bacteria were
investigated only 2 out of the 4 had a significant decrease in relative
abundance between initial and follow up for the carcinoma group and 0
out of the 4 were significant for the adenoma group. This data suggests
that the changes that may be important in the bacterial microbiome after
surgery are not necessarily any one specific bacterium, a clear
depression of CRC related bacteria, or any addition of new or depressed
bacterium. Instead it may be due to a shift of the bacterial community
as a whole between existing members of the host's microbiome.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or before samples in order to confirm that the community was what was
responding or changing due to surgery. We found that the OTUs that made
up the most important variables to the model overwhelmingly belonged to
comensal bacteria. With only the lesion model having a single OTU from a
previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). Using only these important OTUs and FIT both models
(lesion and before sample) had a significantly decreased positive
probability of either lesion or being an initial sample on follow up
{[}Figure 4 \& S4{]}. Further confirmation of the importance of the
changes of comensal bacteria to these classifications was that a total
of 14 OTUs were common to both models and the vast majority belonged to
regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occured. First, one potential explanation is that even with surgery
and a shift of the bacterial community these specific bacteria still
persisted within the colon. Second, the bacteria even if they were
reduced in relative abundance they were still present within the gut.
Third, is that they were not present in enough individuals to be able to
classify those with and without disease with a high degree of accuracy.
Fourth, is that it is possible that our Random Forest models were able
to get the same information from measures such as FIT or other OTUs.
Finally, it is also possible that all three of these potential
explanations could have played a role. Regardless, our observations
would suggest that an individual's resident bacteria have a large role
to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of a lesion after
surgery {[}Figure 4{]}. It should be noted that our study does not argue
against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that the models do not utilize these
specific bacteria for classification purposes (lesion or before sample).
In fact, it is possible that these CRC associated bacteria are important
in the transition from adenoma to carcinoma and would be one explanation
as to why in our data we not only see high initial relative abundances,
in certain individuals, but also large decreases in relative abundance
in those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model to have kept a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the stool DNA test
{[}20{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of overfitting or having a model
that may not be immediately extrapolateable to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur.

Another interesting outcome was that Within figure 3 the before sample
model showed better test AUC results then the training set AUC. This may
have occured because the training AUC that was determined from 20
repeated 10 fold cross validation removed samples at random and did not
take into account that they were matched samples. Another potential
reason is that the model itself may be overfit since the total number of
samples was not that large. However, the lesion model did not suffer
from these discrepencies and similar conclusions can be drawn solely
from this model. Regardless, further independent studies will need to be
carried out to verify our findings since not only are we dealing with
feces, which could be very different than the communities present on the
actual tissue, but also are dealing with correlations that may not be
representative of the true pathogensis of disease.
